---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
--- 


## Loading and preprocessing the data
The original data is provided in a zipped file. To review the data first unzip the file (activity.zip) R and look at the data.
```{r}
# Check if file already exists, else unzip the file
if(!file.exists("activity.csv"))
    unzip(zipfile = "activity.zip")

# Read CSV file
activity_data <- read.csv(file = "activity.csv",na.strings = "NA", stringsAsFactors = FALSE )

# View the data
head(activity_data)
tail(activity_data)
```


## What is mean total number of steps taken per day?
```{r, message=FALSE}
# Use dplyr package
require(dplyr)

# Table of total steps per day
spd <- tbl_df(activity_data) %>% group_by(date) %>% summarize(s=sum(steps,na.rm = TRUE))

# Calc mean of total steps per day
spdm <- mean(spd$s)

# Calc median of total steps per day
spdmd <- median(spd$s)

# Create histogram
hist(spd$s, col="green", main="Histogram of total number of steps taken each day", xlab = "Total steps")

# Add mean
abline(v=spdm, col="magenta", lwd=4)

# Add median
abline(v=spdmd, lwd=4, col="red")

# Add legend
legend("topright", lwd=4, col=c("magenta", "red"), legend=c("Mean","Median"))

# Mean
spdm

# Median
spdmd
```

## What is the average daily activity pattern?
```{r}
# Prepare data by 5-min intervals
spi <- tbl_df(activity_data) %>% group_by(interval) %>% summarize(ave=mean(steps,na.rm = TRUE))

# Create plot
plot(x=spi$interval, y=spi$ave, type="l", xlab = "5-Minute Interval", ylab="Average steps", main="Average daily activity pattern")

# Order data for max average steps
spi <- spi[order(-spi$a),]

# Extract top row
max_spi <- spi[1,]

# Add top activity interval to plot
abline(v=max_spi$interval,b=max_spi$ave,lty=2, lwd=1, col="red")

# Add legend
legend("topright", lwd=1, lty=2, col=c("red"), legend=c(paste("Max activity 5-min interval = ", max_spi$interval)))

# Print max 5-min interval
max_spi$interval
```

## Imputing missing values
### Calculate number of NA rows.
```{r}
isNA  <- is.na(activity_data$steps)

# Number of missing values
length(isNA[isNA==TRUE])
```

### Create new copy of the activity data and update NA values with average 5 minute interval value
```{r}
activity_data2 <- tbl_df(activity_data)

# Replace NA values 
for( i in seq_len(nrow(activity_data2))) {
    if(is.na(activity_data2[i, 1])) {
        k <- as.integer(activity_data2[i,3])
        val <- spi[spi$interval==k,]$ave
        activity_data2[i, 1] <- val
    }
}
```

```{r echo=FALSE}
remove(i)
remove(k)
remove(val)
```

### Check number of NA rows after update, it should be 0.
```{r}
isNA  <- is.na(activity_data2$steps)

# Number of missing values
length(isNA[isNA==TRUE])
```

### Run analysis on changes to the mean aend median
```{r}
# Table of total steps per day
spd2 <- group_by(activity_data2, date) %>% summarize(s=sum(steps,na.rm = TRUE))

# Calc mean of total steps per day
spdm2 <- mean(spd2$s)

# Calc median of total steps per day
spdmd2 <- median(spd2$s)

# Create histogram
hist(spd2$s, col="green", main="Histogram of total number of steps taken each day", xlab = "Total steps")

# Add mean
abline(v=spdm2, col="magenta", lwd=4, lty=2)

# Add median
abline(v=spdmd2, lwd=1, col="red")

# Add legend
legend("topright", lwd=c(4,1), lty=c(2,1), col=c("magenta", "red"), legend=c("Mean","Median"))

# Mean
spdm2

# Median
spdmd2
```
The mean and median are name same after applying the impute strategy. Both mean and median are larger to previous analysis when NA's were removed from the data set. 


## Are there differences in activity patterns between weekdays and weekends?
To view activity patterns on weekdays and weekends, prepare the data to identify weekdays and weekends.
```{r}
require(lubridate)

# Create blank column for dayType
activity_data2$dayType <- ""

# Assign week day type value
for( i in seq_len(nrow(activity_data2))) {
    wd <- wday(ymd(activity_data2[i, 2]))
    if(wd==0 || wd == 6) 
        activity_data2[i, 4] <- "Weekend"
    else
        activity_data2[i, 4] <- "Weekday"
}
```

Prepare date for plotting.
```{r}
spid <- group_by(activity_data2, interval, dayType) %>% summarize(ave=mean(steps,na.rm = TRUE))
require(ggplot2)
ggplot(spid, aes(interval, ave)) + geom_line() + facet_wrap(~dayType, nrow = 2, ncol=1) + labs(x="5-Minute Interval") + labs(y = "Average number of steps") + labs(title="Activity patterns on weekdays and weekends")
```

As seen in the plots it is clear that activity during the weekends is higher.
